

\section*{AUTHOR CONTRIBUTIONS}

JME wrote \S \ref{sec:positions}, \S \ref{sec:graphalignment}, \S \ref{sec:graphmapping}, and \S \ref{sec:variantcalling}, made Table \ref{table:mappers}, and revised the paper;
AMN wrote \S \ref{sec:viz}, contributed to \S \ref{sec:graphmapping}, and revised the paper;
JAS wrote \S \ref{sec:haplotypes} and \S \ref{sec:functionalgenomics}, and contributed to section \ref{sec:graphmapping};
SH made Table \ref{table:Visualization_Features} and contributed to \S \ref{sec:constructionWGA}, \S \ref{sec:viz}, and Figure \ref{fig:visualization}.
AG made Figure \ref{fig:models};
GH contributed to \S \ref{sec:variantcalling};
XC contributed to \S \ref{sec:graphmapping};
JDS contributed to Table \ref{table:Visualization_Features} and contributed significantly to \S \ref{sec:viz};
RR contributed to \S \ref{sec:building};
JE contributed to \S \ref{sec:variantcalling};
MR contributed to \S \ref{sec:applications};
SG contributed to \S \ref{sec:haplotypes};
BP provided guidance and opinion;
TM contributed to \S \ref{sec:intro}, \S \ref{sec:resequencing} and \S \ref{sec:graphalignment};
JS wrote \S \ref{sec:indexing};
EG organized the work, wrote \S \ref{sec:intro}, \S \ref{sec:resequencing}, \S \ref{sec:models}, \S \ref{sec:formats}, \S \ref{sec:discussion}, made Figures \ref{fig:visualization} and \ref{fig:allelebalance}, and revised the paper.

%Disclosure
\section*{DISCLOSURE STATEMENT}
The authors are not aware of any affiliations, memberships, funding, or financial holdings that
might be perceived as affecting the objectivity of this review. 

% Acknowledgements
\section*{ACKNOWLEDGMENTS}
Research reported in this publication was supported the National Institutes of Health under Award Numbers U54HG007990, U01HL137183 and 2U41HG007234.
Its contents are solely the responsibility of the authors and do not necessarily represent the official views of National Institute of Health.
The research was also made possible by the generous financial support of the W.M. Keck Foundation (DT06172015).
T.M.\ acknowledges funding from the German Federal Ministry for Research and Education (BMBF 031L0184).
The work of J.A.S. was supported by the Carlsberg Foundation.
S.H. acknowledges funding from the central innovation programme (ZIM) for SMEs of the Federal Ministry for Economic Affairs and Energy of Germany.
J.D.S. thanks BBSRC for funding BB/S004661/1.
We would also like to thank all the attendees of the joint NCBI/UCSC Pangenomics Hackathon which took place at UCSC in the spring of 2019 and which spurred many conversations that contributed to this review.
